\chapter{Conclusion and Future Work}
\label{chapter:conclusion}

In this research we developed a POMDP formulation for the GRN control problem and presented an algorithm for
solving the problem efficiently by decomposing it to subproblems that exploit the problem structure. In fact,
our approach will be beneficial and influential for more realistic GRN interference mechanisms. By
considering partial observability, the problem is expressed in a difficult setting and advancements in the
settings might also lead to more efficient and successful solutions of the problem.% for simpler settings too.
After achieving the successful and encouraging results reported in this paper, we are currently concentrating
on advancing this work along two tracks. The setting proposed for the GRN control problem has definitely room
for improvements. Different kinds of gene expression data sampling techniques have been developed recently.
The gene expression data analysis part can be improved, for using the results more effectively, by employing
different gene expression data sampling techniques. Also in this work we did not explore possible
correlations or similarities between the GRN and the subproblems. By examining such relationship, it might be
possible to guide the decomposition and have more effective decomposition mechanisms that lead to greater
improvements in computational cost of the solution. The proposed POMDP formulation of the gene expression control problem
assumes that genes influence each other linearly. It is possible to relax this assumption by introducing a
more general transition function. Finally, it is worth mentioning that this method has been
actually developed for a generic POMDP problem and then adapted to the GRN control problem in hand. Gene
expression data has proved to be valuable resource for guiding the decomposition process. Similar problems,
where data about the problem is available, can be tackled with our approach. However, for a more general
POMDP decomposition method, we need to analyze the role of the gene expression data and a couple of other
components; we should also formalize a more general guidance mechanism.

