\documentclass[11pt,oneandhalf,chaparabic]{metu}

\usepackage{appendix}
%\newcommand{\HRule}{\rule{\linewidth}{0.5mm}}
%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{definition}[theorem]{Definition}
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{observation}[theorem]{Observation}
%\newtheorem{example}[theorem]{Example}

%\newcommand{\eod}{$\quad\triangleleft$}

%\newenvironment{defn}{\begin{definition}}{\hfill$\blacksquare$\end{definition}}
%\newenvironment{defn*}{\begin{definition}}{\end{definition}}
%\newenvironment{proof}{\noindent{\bf Proof.}}{\hfill$\blacksquare$}

% mypackages
\usepackage{tikz}
\usetikzlibrary{positioning,arrows,backgrounds,fit,matrix}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[boxed,figure]{algorithm2e}
\usepackage[all]{xy}
\usepackage{paralist}
\usepackage{multirow}

\newcommand\fBold[1]{\textbf #1\relax}

\author{Utku Erdoğdu}
\authoruppercase{UTKU ERDOĞDU}
\graduateschool{Natural and Applied Sciences}
\shortdegree{Ph.D.}
%\shortdegree{M.S.}
\director{Prof. Dr. Canan Özgen}
\headofdept{Prof. Dr. Adnan Yazıcı}

\title{EFFICIENT PARTIALLY OBSERVABLE MARKOV DECISION PROCESS BASED FORMULATiON OF GENE REGULATORY NETWORK CONTROL PROBLEM}
\degree{Doctor of Philosophy}
%\degree{Master of Science}
\department{Computer Engineering}
\date{September 2011}
\supervisor{Prof. Dr. Faruk Polat}
\departmentofsupervisor{Computer Engineering Department, METU}
\cosupervisor{Prof. Dr. Reda Alhajj}
\departmentofcosupervisor{Computer Science Department, University of Calgary}


\turkishtitle{GEN AĞLARININ KISMİ GÖZLEMLENEBİLİR MARKOV KARAR SÜREÇLERİ İLE MODELLENEREK ETKİN OLARAK KONTROLÜ}
\turkishdegree{Doktora}
%\turkishdegree{Yüksek Lisans}
\turkishdepartment{Bilgisayar Mühendisliği Bölümü}
\turkishdate{Eylül 2011}
\turkishsupervisor{Prof. Dr. Faruk Polat}
\turkishcosupervisor{Prof. Dr. Reda Alhajj}


\keywords{Gene Regulatory Networks, Partially Observable Markov Decision Process, Control of GRN}
\anahtarklm{Gen Düzenleyici Ağlar, Kısmi Gözlemlenebilir Markov Karar Süreçleri, GDA'ların Kontrolü}

\committeememberi{Prof. Dr. Varol Akman}
\affiliationi{Computer Engineering, Bilkent University}
\committeememberii{Prof. Dr. Faruk Polat}
\affiliationii{Computer Engineering, METU}
\committeememberiii{Assoc. Prof. Dr. Tolga Can}
\affiliationiii{Computer Engineering, METU}
\committeememberiv{Assoc. Prof. Dr. Halit Oğuztüzün}
\affiliationiv{Computer Engineering, METU}
\committeememberv{Assist. Prof. Dr. Mehmet Tan}
\affiliationv{Computer Engineering, TOBB ETÜ}

\begin{document}
\begin{preliminaries}
\maketitle
\makeapproval
\plagiarism
\setlength{\parindent}{0em}
\setlength{\parskip}{10pt}

\begin{abstract} \oneandhalfspacing
The need to analyze and closely study the genes related mechanism motivated for the research on the 
modeling and control of gene regulatory networks (GRN). Different approaches exist to model GRNs; they 
are mostly simulated as mathematical models that represent relationships between genes. The GRN control 
problem has been studied mostly with the aid of probabilistic boolean networks, and corresponding control 
policies have been devised. Though turns into a more challenging problem, we argue that partial observability 
would be a more natural and realistic method for handling the control of GRNs. Partial observability 
is a fundamental aspect of the problem; it is mostly ignored and substituted by assumption that states 
of GRN are known precisely, prescribed as full observability. On the other hand, current works addressing 
partially observability focus on formulating algorithms for the finite horizon GRN control problem. 
So, in this work we explore the feasibility of realizing the problem in a partially observable 
setting, mainly with Partially Observable Markov Decision Processes (POMDP). The method proposed in 
this work is a POMDP formulation for the infinite horizon version of the problem. This formulation 
first decomposes the problem by isolating different unrelated parts of the problem, and then makes 
use of existing POMDP solvers to solve the obtained subproblems; the final outcome is a control mechanism 
for the main problem. The reported test results using both synthetic and real GRNs are promising in 
demonstrating the applicability, effectiveness and efficiency of the proposed approach.
\end{abstract}

\begin{oz} \oneandhalfspacing
Genlerin çalışma prensiplerini inceleme gereksinimi gen düzenleyici ağların (GDA) modellenmesi ve kontrolü
üzerine bilimsel çalışmalar yapılmasına yol açmıştır. GDA'ları modellemek için değişik yaklaşımlar mevcuttur
ve bu yaklaşımların çoğu genler arasındaki ilişkileri matematiksel modeller vasıtasıyla modellemektedir.
GDA kontrol problemi en çok olasılıksal boole ağları yardımıyla çalışılmaktadır ve bu model yardımıyla
kontrol problemlerine çözümler üretilmektedir. Problemi daha zorlaştırmasına rağmen, GDA kontrol problemlerinin
daha doğal ve gerçekçi çözülebilmesi için kısmi gözlemlenebilirliğin önerilmesi gerektiğini savunuyoruz. 
Kısmi gözlemlenebilirlik bu problemin temel bir bileşeni olmasına rağmen çoğunlukla gözardı edilmiş ve
problemin çözümünde GDA'nın tüm durumlarının mükemmel olarak bilinebileceği varsayımı yapılmıştır, yani
problem tam gözlemlenebilir kabul edilmiştir. Bir yandan da literatürdeki kısmi gözlemlenebilirliği
dikkate alan yöntemler sınırlı adımdan oluşan bir problem tanımı ile GDA kontrol problemini çözen algoritmalar
üretmeye çalışmaktadır. Bu çalışmada problemin kısmi gözlemlenebilir bir kurgu ile tanımlanması üzerinde
çalışılmakta ve Kısmi Gözlemlenebilir Markov Karar Süreçleri (POMDP) bu kurguda kullanılmaktadır. Bu çalışmada
önerilen metot problemin sonsuz adımdan oluşan bir halinin POMDP ile tanımlanmasıdır. Bu tanımlama
öncelikle problemin birbirinden bağımsız kısımlarını soyutlamakta ve daha sonra elde edilen alt-problemleri
varolan bir POMDP çözücü yardımıyla çözmektedir; sonuçta ana problem için bir çözüm elde edilmektedir. 
Sunulan test sonuçları önerilen metodun hem doğal hem de yapay GDA'lar için uygulanabilirliğini, etkisini
ve başarımını göstermek konusunda başarılıdır.

\end{oz} 

\dedication{\textit{Dedication Goes Here}}

\setlength{\parindent}{0em}
\setlength{\parskip}{10pt}

\begin{acknowledgments} \oneandhalfspacing
Acknowledgements Goes Here
\end{acknowledgments}
\setlength{\parindent}{0em}
\setlength{\parskip}{3pt}

\tableofcontents
\listoftables
\listoffigures
\end{preliminaries}
   
\setlength{\parindent}{0em}
\setlength{\parskip}{10pt}

\input{introduction/introduction.tex}
\newpage

\input{relatedwork/relatedwork.tex}
\newpage

\input{background/background.tex}
\newpage

\input{formulation/formulation.tex}
\newpage

\input{dataanalysis/dataanalysis.tex}
\newpage

\input{decomposition/decomposition.tex}
\newpage

\input{experiments/experiments.tex}
\newpage

\input{conclusion/conclusion.tex}
\newpage

\bibliography{thesis_references}{}
\bibliographystyle{plain}
%\appendix
%\input{appendix1/appendix1.tex}
\end{document}
